```{r}
library(tidyverse)        # For Generic data wrangling
library(data.table)       # For efficient data wrangling
library(lubridate)        # For working with date times
library(naniar)           # View missing patterns

setwd("/home/jeffrw/BIOSTAT629/main_project/preprocessing")
```



# Load preprocessed datasets
```{r}
# Load datasets from TSV files
survey_data <- read_tsv("../data/intermediate/survey_data.tsv", show_col_types = F)
daily_data <- read_tsv("../data/intermediate/daily_data.tsv", show_col_types = F)
daily_data_imputed <- read_tsv("../data/intermediate/daily_data_imputed.tsv", show_col_types = F)
sleep_data <- read_tsv("../data/intermediate/sleep_data.tsv", show_col_types = F)
sleep_data_imputed <- read_tsv("../data/intermediate/imputed_sleep_data.tsv", show_col_types = F)
```

# Find common PatientIDs across all datasets
```{r}
common_patient_ids <- Reduce(intersect, list(
  survey_data$PatientID, 
  daily_data$PatientID, 
  daily_data_imputed$PatientID, 
  sleep_data$PatientID, 
  sleep_data_imputed$PatientID
))

# Filter all datasets to keep only common PatientIDs
survey_data <- survey_data %>% filter(PatientID %in% common_patient_ids)
daily_data <- daily_data %>% filter(PatientID %in% common_patient_ids)
daily_data_imputed <- daily_data_imputed %>% filter(PatientID %in% common_patient_ids)
sleep_data <- sleep_data %>% filter(PatientID %in% common_patient_ids)
sleep_data_imputed <- sleep_data_imputed %>% filter(PatientID %in% common_patient_ids)


common_patient_ids %>% length()
````

# Filter Daily & Sleep Data Based on Survey Start & End Dates
```{r}
# Function to filter data based on survey start and end dates
filter_dates_within_survey <- function(data, survey_data) {
  data_filtered <- data %>%
    inner_join(survey_data %>% select(PatientID, start_date, end_date), by = "PatientID") %>%
    filter(Date >= start_date & Date <= end_date) %>%
    select(-start_date, -end_date)  # Remove extra columns after filtering
  
  return(data_filtered)
}

# Apply function to filter daily and sleep datasets
daily_data <- filter_dates_within_survey(daily_data, survey_data)
daily_data_imputed <- filter_dates_within_survey(daily_data_imputed, survey_data)
sleep_data <- filter_dates_within_survey(sleep_data, survey_data)
sleep_data_imputed <- filter_dates_within_survey(sleep_data_imputed, survey_data)
```

# Helper functions
```{r}
# Function to compute and plot CDF using ecdf()
plot_cdf <- function(data, dataset_name) {
  date_counts <- data %>%
    group_by(PatientID) %>%
    summarize(num_dates = n()) 
  
  # Compute the empirical CDF
  ecdf_func <- ecdf(date_counts$num_dates)
  
  # Create CDF plot (1 - ECDF for "at least" interpretation)
  ggplot(data.frame(x = unique(date_counts$num_dates)), aes(x = x)) +
    stat_function(fun = function(x) 1 - ecdf_func(x), geom = "line", color = "blue", size = 1) +
    theme_minimal() +
    labs(
      title = paste("Cumulative CDF of Recorded Dates in", dataset_name),
      x = "Number of Recorded Dates",
      y = "Proportion of Subjects (≥ X Dates)"
    )
}

# Function to check for non-consecutive dates
plot_non_consecutive_histogram <- function(data, dataset_name) {
  data %>%
    arrange(PatientID, Date) %>%
    group_by(PatientID) %>%
    summarize(non_consecutive = sum(diff(as.Date(Date)) > 1)) %>%
    ggplot(aes(x = non_consecutive)) +
    geom_histogram(binwidth = 1, fill = "blue", alpha = 0.7) +
    theme_minimal() +
    labs(
      title = paste("Histogram of Non-Consecutive Dates in", dataset_name),
      x = "Number of Skipped Days",
      y = "Number of Subjects"
    )
}


daily_data_imputed
```


```{r}
plot_cdf(daily_data, "Daily Data")
plot_cdf(daily_data_imputed, "Daily Imputed Data")
plot_cdf(sleep_data, "Sleep Data")
plot_cdf(sleep_data_imputed, "Sleep Imputed Data")

```
```{r}
plot_non_consecutive_histogram(daily_data, "Daily Data")
plot_non_consecutive_histogram(daily_data_imputed, "Daily Imputed Data")
plot_non_consecutive_histogram(sleep_data, "Sleep Data")
plot_non_consecutive_histogram(sleep_data_imputed, "Sleep Imputed Data")

```

```{r}
# Perform an outer join on PatientID and Date
merged_data <- full_join(daily_data_imputed, sleep_data_imputed, by = c("PatientID", "Date"))

# Sort by PatientID and Date
merged_data <- merged_data %>% arrange(PatientID, Date)

# Perform a left join to bring in survey data
merged_data <- left_join(merged_data, survey_data, by = "PatientID")


```

# Only select subjects with at least 21 days worth of recrods
```{r}
# Function to filter patients with at least 21 entries
filter_patients_with_min_entries <- function(data, min_entries = 21) {
  data %>%
    group_by(PatientID) %>%
    filter(n() >= min_entries) %>%
    ungroup()  # Ensure the result is ungrouped
}

# Apply the function to filter the dataset
filtered_data <- filter_patients_with_min_entries(merged_data)

```

# INFO EXTRACTION, REMOVE LATTER
```{r}
merged_data %>%
  group_by(PatientID) %>%
  summarize(totals = n()) %>%
  pull(totals) %>%
  table() %>%
  prop.table() %>%
  cumsum()

```
```{r}
# # Calculate date differences per subject
# date_diffs <- merged_data %>%
#   arrange(PatientID, Date) %>%
#   group_by(PatientID) %>%
#   mutate(date_diff = as.numeric(Date - lag(Date))) %>%
#   filter(!is.na(date_diff))  # Remove first entry per patient since it has no diff
# 
# overall_avg <- mean(date_diffs$date_diff)
# overall_sd <- sd(date_diffs$date_diff)
# 
# cat("Overall average date difference:", overall_avg, "\n")
# cat("Overall standard deviation:", overall_sd, "\n")

```



# Check number of samples
```{r}
filtered_data %>% pull(PatientID) %>% unique() %>% length()
```

# Function to view missing rows that are not from sleep data
```{r}
missing_data_rows <- function(df, sleep_data) {
  # Define the excluded columns dynamically from sleep_data
  exclude_columns <- sleep_data %>% colnames()
  
  # Select columns that are NOT in the exclude list
  relevant_data <- df[, !(colnames(df) %in% exclude_columns)]
  
  # Identify rows where all missing values are ONLY in the excluded columns
  missing_in_excluded_only <- apply(df, 1, function(row) {
    all(is.na(row[colnames(df) %in% exclude_columns])) && 
    all(!is.na(row[!(colnames(df) %in% exclude_columns)]))
  })
  
  # Return the matching rows
  return(df[missing_in_excluded_only, ])
}

# Only missing data in sleep_data
missing_data_rows(filtered_data, sleep_data)
```

# fill NA with 0, and add skipped indicator
```{r}
update_skipped_and_fill_na <- function(df) {
  df <- df %>%
    mutate(skipped = ifelse(rowSums(is.na(.)) > 0, 1, 0)) %>% # Set skipped to 1 if any NA in row
    mutate(across(everything(), ~ replace_na(., 0))) # Replace all NAs with 0
  
  return(df)
}


filtered_data_filled <- update_skipped_and_fill_na(filtered_data)
```

# get timepoints
```{r}
timepoints <- function(df) {
  df %>%
    group_by(PatientID) %>%
    arrange(Date, .by_group = TRUE) %>%  # Ensure data is sorted by Date within each PatientID
    mutate(timepoints = as.integer(difftime(Date, min(Date), units = "days"))) %>%
    ungroup() %>% 
    mutate(timepoints = timepoints + 1)# Remove grouping for final output
}

# Apply the function to filtered_data_filled
filtered_data_cleaned <- timepoints(filtered_data_filled)

```
#################### OLD CODE HERE ####################


# Ensure non-imputed data and survey data have same date/id as the filtered set
```{r}
# Compute common PatientIDs across all datasets
common_patient_ids <- Reduce(intersect, list(
  unique(filtered_data_cleaned$PatientID),
  unique(daily_data$PatientID),
  unique(sleep_data$PatientID),
  unique(survey_data$PatientID)
))

# Extract final PatientIDs and Dates from filtered_data_cleaned for the common patients
final_patient_dates <- filtered_data_cleaned %>%
  filter(PatientID %in% common_patient_ids) %>%
  select(PatientID, Date) %>%
  distinct()

# Filter daily_data and sleep_data using an inner join on PatientID and Date
daily_data_cleaned <- daily_data %>%
  inner_join(final_patient_dates, by = c("PatientID", "Date"))

sleep_data_cleaned <- sleep_data %>%
  inner_join(final_patient_dates, by = c("PatientID", "Date"))

# Filter survey_data to keep only the common PatientIDs
survey_data_cleaned <- survey_data %>%
  filter(PatientID %in% common_patient_ids)

# Filter filtered_data_cleaned to keep only the common PatientIDs
filtered_data_cleaned <- filtered_data_cleaned %>%
  filter(PatientID %in% common_patient_ids)



# Define the columns to remove
cols_to_remove <- c(
  # Specified Heart variables to remove
  "HeartRateZoneOutOfRangeMin",
  "HeartRateZonePeakMax",
  "HeartRateZoneFatBurnMin",
  "HeartRateZoneCardioMin",
  "HeartRateZonePeakMin",
  
  # Remove non-tracker versions when a Tracker version exists
  "ActivityCalories",
  "Calories",
  "Distance",
  "MinutesFairlyActive",
  "MinutesLightlyActive",
  "MinutesSedentary",
  "MinutesVeryActive",
  "Steps"
)

daily_data_cleaned %>% colnames()

# Need to remove above columns from daily, since we removed them based on imputation before(see fibit_daily_preprocessing)
daily_data_cleaned <- daily_data_cleaned %>% select(-all_of(cols_to_remove))

```

# Filter out unneeded columns for modeling
```{r}
filtered_data_cleaned %>% colnames()
```
```{r}
filtered_data_cleaned <- filtered_data_cleaned %>% select(-c(Date, end_date, start_date, Weeks_Difference, Baseline_Date))
```



# Save cleaned files
```{r}
# Save all cleaned datasets
write_tsv(daily_data_cleaned, "../data/cleaned/daily_data_cleaned.tsv")
write_tsv(sleep_data_cleaned, "../data/cleaned/sleep_data_cleaned.tsv")
write_tsv(survey_data_cleaned, "../data/cleaned/survey_data_cleaned.tsv")
write_tsv(filtered_data_cleaned, "../data/cleaned/combined_data_cleaned.tsv")  # Save final combined dataset


```


